Parsing Postfix log files

### 2009-02-10 doc/logparser-conclusion.tex
Dealt with some XXX items.

The division of the parser into rules, actions, and framework is unusual
because rules are separated so completely from the actions and framework;
although parsers are often divided (the combination of \texttt{lex} and
\texttt{yacc}~\cite{lex-and-yacc-book} being a common example), the parts
are usually quite internally interdependent, and will be combined into a
complete parser by the compilation process; in contrast \parsername{} keeps
the rules and actions separate until the parser runs.  The actions and
framework are not as completely separated, as the actions depend on
services provided by the framework; however the actions and framework are
not tightly integrated: it would be possible, with some work, to completely
separate the two.  XXX I SHOULD DO THAT\@.

% XXX THIS SHOULD PROBABLY BE REMOVED\@.
%The emergent behaviour (\sectionref{Emergent behaviour}) exhibited by the
%rules and actions is also interesting, and is discussed after the flow
%chart (\sectionref{flow-chart}) and explanation of the paths through the
%parser.  This emergent behaviour greatly eases the process of adding new
%actions (\sectionref{adding new actions}), as the actions do not have to
%be inserted into an explicit flow of control; new actions will naturally
%find their place in the paths through the parser.

>>>>

The division of the parser into rules, actions, and framework is unusual
because rules are separated so completely from the actions and framework;
although parsers are often divided (the combination of \texttt{lex} and
\texttt{yacc}~\cite{lex-and-yacc-book} being a common example), the parts
are usually quite internally interdependent, and will be combined into a
complete parser by the compilation process; in contrast \parsername{} keeps
the rules and actions separate until the parser runs.  The actions and
framework are not as completely separated, because the actions depend on
services provided by the framework; however the actions and framework are
not tightly integrated: it would be possible, with some work, to completely
separate the two.

### 2009-02-10 doc/logparser-implementation.tex
Dealt with some XXX items.

    \item [regex] XXX THIS NEEDS TO BE UPDATED ONCE AUTOMATIC DATA
        EXTRACTION IS FINISHED\@.  The regex to match the log line
        against.  The regex will first have several keywords expanded:
        this simplifies reading and writing rules; avoids needless
        repetition of complex regex components; allows the components to
        be corrected and/or improved in one location; and makes each
        regex largely self-documenting.  For efficiency the keywords are
        expanded and every rule's regex is compiled before attempting to
        parse the log file --- otherwise each regex would be recompiled
        each time it was used, resulting in a large, data dependent
        slowdown.  Efficiency concerns are discussed in \sectionref{parser
        efficiency}, with caching compiled regexes covered in
        \sectionref{Caching compiled regexes}.

    \item [result\_data] Sometimes rules need to save a result that is not
        present in the log line: e.g.\ setting \texttt{smtp\_code} when
        mail is accepted.  The format is: \newline{} \tab{} \texttt{XXX
        UPDATE EXAMPLE\@; smtp\_code = 550; \newline{} recipient = 2,
        sender = 4;} \newline{} i.e.\ semi-colon or comma separated
        assignment statements, with the variable name on the left and the
        value on the right hand side.\footnote{Commas and semi-colons
        cannot be escaped and thus cannot be used.  This feature is
        intended for use with small amounts of data rather than large
        amounts in any one rule, so dealing with escape sequences was
        deemed unnecessary.}  Any field in the results table can be set in
        this way; additionally \texttt{child} and \texttt{pid} are accepted
        and used respectively by the \texttt{TRACK}, \texttt{BOUNCE}, and
        \texttt{SMTPD\_DIED} actions, but do not cause data to be entered
        in the database.

    \item [connection\_data] The connections table equivalent of
        \texttt{result\_data}.

@@

    \item [EXPIRED] The mail has been in the queue for five
        days\footnotemark[\value{footnote}] XXX CHECK IF THIS
        FOOTNOTE WORKS without successful delivery.  A bounce mail
        will be generated and sent to the sender address.

@@

\textbf{Field}      & \textbf{Value}                                    \\
name                & Unknown sender domain                             \\
description         & We do not accept mail from unknown domains        \\
restriction\_name   & reject\_unknown\_sender\_domain                   \\
postfix\_action     & REJECTED                                          \\
program             & \daemon{smtpd}                                    \\
XXX                 & UPDATE REGEX FOR AUTOMATIC SAVING                 \\
regex               & \verb!^__RESTRICTION_START__ <(__SENDER__)>: !    \\
                    & \verb!Sender address rejected: Domain not found;! \\
                    & \verb!from=<\5> to=<(__RECIPIENT__)> !            \\
                    & \verb!proto=E?SMTP helo=<(__HELO__)>$!            \\
result\_data        &                                                   \\
connection\_data    &                                                   \\
action              & DELIVERY\_REJECTED                                \\
hits                & 0                                                 \\
hits\_total         & 0                                                 \\
priority            & 0                                                 \\
%cluster\_group      & 400                                               \\

@@

    \item [matches] The fields in the line captured by the rule's regex.

@@

@@

\Figureref{flow chart image} shows the paths the data representing a mail
can take through the parser algorithm.  The flow chart covers the most
common paths only, the complications described in
\sectionref{complications} are excluded for the sake of clarity.  XXX WHERE
ARE THE PATHS DESCRIBED\@?

@@

\clearpage

\subsection{Mail enters the system}

@@

XXX SHOULD THESE BE SEPARATE SUBSUBSECTIONS\@?

Everything starts off with a mail entering the system, whether by local
submission via \daemon{postdrop}, by \gls{SMTP}, by re-injection due to
forwarding, or internally generated by Postfix.  Local submission is the
simplest case: a queueid is assigned immediately and the sender address is
logged (action: PICKUP\@; flowchart:~2).

\gls{SMTP} is more complicated:

@@

Re-injection due to forwarding sadly lacks explicit log lines of its
own; it is explained in \sectionref{Re-injected mails}.

\subsection{Mail delivery}

@@

\subsection{Emergent behaviour}

\label{Emergent behaviour}

XXX DUNNO IF I SHOULD INCLUDE THIS SECTION\@; AT BEST IT NEEDS TO BE
UPDATED, AT WORST IT IS RUBBISH\@.

The rules and actions exhibit emergent behaviour that is far more
complicated than the behaviour explicitly encoded in the algorithm.  The
top-level parser could be written in pseudo-code like so (with indentation
level indicating flow control):

\begin{verbatim}
for each line in the input files:
    for each rule defined by the user:
        if this rule matches the input line:
            perform the action specified by the rule
            skip the remaining rules
            process the next input line
    warn the user that the input line was not parsed
\end{verbatim}

Yet despite the brevity of the algorithm above the (simplified) flow chart
in \sectionref{flow-chart} contains:

    \item Twenty one states (not including explicit branches).

    \item Three entry points (1, 2, 3).

    \item Five exit points (8, 14, 17, 20, 25).

    \item Five loops (4--4, 4--5--4, 5--5, 9--9, 9--10--9).

    \item Four explicit branches (7, 13, 15, 18); these are decisions taken
        by an action, determining what will happen next to a mail.

    \item Four implicit branches, where the transition is determined by the
        next log line that is processed for that mail, e.g.\ state 1 can
        transition to either state 4 or state 5.  (1--\{4,5\},
        4--\{4,5,6\}, 5--\{4,5,6\}, 9--\{9,10,11,12\})

Several more states and corresponding branches are omitted for the sake of
clarity (these extra states are caused by solving the complications
described in \sectionref{complications}).

The actions introduce some of the complexity seen in the flow chart: all
explicit branches and the transitions between some stages (19--20,
21--22--23--24) in the second half of the flow chart are encoded within the
actions; the transitions between the other stages (16--17, 24--25) in the
second half arise from the interaction between actions

However the transitions between stages in the first half of the flow chart
(\figureref{flow chart image}) are not encoded anywhere in
the parser; the transitions are determined by the ordering of log lines in
the log file.  The complexity of the first 12 stages in the flow chart
emerges from the interaction of simple rules and actions, easing the
process of adding new actions (described in \sectionref{adding new
actions}), as there is no requirement to explicitly insert the action into
an algorithmic version of the flow chart.\footnote{The solution to the
complication in \sectionref{out of order log lines} (out of order log
lines) requires a list of the acceptable combinations of Postfix programs a
mail can pass through; however this does not correspond to the
\textit{parser actions\/} a mail must pass through.  Before that
complication was overcome, i.e.\ for the first half of the development of
this parser, there was nothing in the parser explicitly encoding the states
or paths a mail could take; no other component in the parser has any need
for this flow of data to be specified.}

@@

XXX DO I NEED THIS LIST\@? IMPROVE HOW IT LOOKS IF I KEEP IT\@.

Different delivery methods result in different log lines:

    \item [Re-injected for forwarding:] forwarded as 56F5B43FD

    \item [Delivered to a remote \gls{SMTP} server:] 250 Ok: queued as
        BD07D3C49

    \item [Local delivery:] delivered to command: /usr/bin/procmail

@@

subsubsection{Tracking re-injected mail}

The crux of the problem is that re-injected mails appear in the log files
without explicit logging indicating their source.  There are two implicit
indications:

@@

    \item The indicator which more commonly introduces re-injection is when
        \daemon{qmgr} selects a mail with a previously unseen queueid for
        delivery (action: MAIL\_QUEUED\@; flowchart:~3), in which case a
        new data structure will be created for that mail.  The mail will be
        flagged as having unknown origins; this flag should be subsequently
        cleared once the origin has been established.  This may also be an
        indicator that the mail is a bounce notification, see
        \sectionref{identifying-bounce-notifications} for details.

    \item Local delivery re-injects the mail and logs a relayed delivery
        rather than delivering directly to a mailbox or program as it
        usually would (action: TRACK\@; flowchart:~10).\footnote{Relayed
        delivery is performed by the \gls{SMTP} client; local delivery
        means local to the server, i.e.\ an address the server is final
        destination for.} In this case the mail may already have been
        created (described above) and the unknown origin flag will be
        cleared; if not a new data structure will be created.  In both
        cases the re-injected mail is marked as a child of the original
        mail.  The log line in question is:

        \texttt{3FF7C4317: to=<username@example.com>, relay=local,
        \newline{} delay=0, status=sent (forwarded as 56F5B43FD)}

        This second indicator always occurs for re-injected mail but
        typically occurs after the first indicator explained above.  This
        is required to tie the parent and child mails together and so is
        central to the process of tracking re-injected mails.

@@

    \item If the mail is of unknown origin it is assumed to be a child mail
        whose parent has not yet been identified (action: COMMIT\@;
        flowchart:~15).  Mark the mail as ready for entry in the database
        (flowchart:~16), and wait for the parent to deal with it
        (flowchart:~17).  The mail should not have subsequent log lines;
        only its parent should refer to it.

    \item If the mail is a child mail then it has already been tracked
        (action: COMMIT\@; flowchart:~18): as with all other mail, the data
        is cleaned up, the child is entered in the database
        (flowchart:~19), then deleted from the state tables.  The child
        mail will be removed from the parent mail's list of children
        (flowchart:~20); if this is the last child and the parent has also
        been entered in the database the parent will be deleted from the
        state tables.

    \item The last alternative is that the mail is a parent mail (action:
        COMMIT\@; flowchart:~21).  Regardless of the state of its children
        its data is cleaned up and entered in the database (flowchart:~22).
        The parent may have children that are waiting to be entered in the
        database (flowchart:~23); each of those children's data is cleaned
        up and entered in the database, then deleted from the state tables.
        The parent may also have outstanding children which are not yet
        delivered (flowchart:~24), in which case the parent must wait for
        those children to be finished with.  As soon as the last child is
        deleted from the state tables the parent will also be finished with
        (flowchart:~25), and deleted from the state tables.

>>>>

    \item [regex] The regex to match the log line against.  The regex will
        have several keywords expanded when the rules are loaded: this
        simplifies reading and writing rules; avoids needless repetition of
        complex regex components; allows the components to be corrected
        and/or improved in one location; enables automatic extraction and
        saving of data; and makes each regex largely self-documenting.  For
        efficiency the keywords are expanded and every rule's regex is
        compiled before attempting to parse the log file --- otherwise each
        regex would be recompiled each time it was used, resulting in a
        large, data dependent slowdown.  Efficiency concerns are discussed
        in \sectionref{parser efficiency}, with caching compiled regexes
        covered in \sectionref{Caching compiled regexes}.

    \item [connection\_data] Sometimes rules need to save data that is not
        present in the log line: e.g.\ setting \texttt{client\_ip} when a
        mail is being delivered to another server.  The format is:
        \newline{} \tab{} \texttt{ client\_hostname = localhost, client\_ip
        = 127.0.0.1} \newline{} i.e.\ semi-colon or comma separated
        assignment statements, with the variable name on the left and the
        value on the right hand side.\footnote{Commas and semi-colons
        cannot be escaped and thus cannot be used.  This feature is
        intended for use with small amounts of data rather than large
        amounts in any one rule, so dealing with escape sequences was
        deemed unnecessary.}  Any field in the connections table can be set
        in this way.

    \item [result\_data] The result table equivalent of
        \texttt{connection\_data}.

@@

@@

\textbf{Field}      & \textbf{Value}                                    \\
name                & Unknown sender domain                             \\
description         & We do not accept mail from unknown domains        \\
restriction\_name   & reject\_unknown\_sender\_domain                   \\
postfix\_action     & REJECTED                                          \\
program             & \daemon{smtpd}                                    \\
regex               & \verb!^__RESTRICTION_START__ <(__SENDER__)>: !    \\
                    & \verb!Sender address rejected: Domain not found;! \\
                    & \verb!from=<\k<sender>> to=<(__RECIPIENT__)> !    \\
                    & \verb!proto=E?SMTP helo=<(__HELO__)>$!            \\
result\_data        &                                                   \\
connection\_data    &                                                   \\
action              & DELIVERY\_REJECTED                                \\
hits                & 0                                                 \\
hits\_total         & 0                                                 \\
priority            & 0                                                 \\
%cluster\_group      & 400                                               \\

@@

    \item [matches] The data extracted from the line by the rule's regex.

@@

\subsection{Postfix actions}

XXX WHY DO I NEED HAVE postfix\_actions AT ALL\@?

THERE IS NEARLY A 1--1 MAPPING BETWEEN action AND postfix\_action.

\begin{tabular}{lll}

1 & BOUNCE & BOUNCED \\
1 & CLONE & ACCEPTED \\
3 & COMMIT & DISCARDED \\
1 & COMMIT & INFO \\
1 & CONNECT & INFO \\
1 & DELETE & DELETED \\
64 & DELIVERY\_REJECTED & REJECTED \\
1 & DISCONNECT & INFO \\
1 & EXPIRY & EXPIRED \\
1 & MAIL\_QUEUED & INFO \\
1 & MAIL\_QUEUED & PROCESSING \\
1 & MAIL\_TOO\_LARGE & DISCARDED \\
1 & PICKUP & INFO \\
2 & POSTFIX\_RELOAD & POSTFIX\_RELOAD \\
13 & SAVE\_DATA & BOUNCED \\
38 & SAVE\_DATA & INFO \\
1 & SAVE\_DATA & REJECTED \\
5 & SAVE\_DATA & SENT \\
2 & SMTPD\_DIED & IGNORED \\
1 & SMTPD\_WATCHDOG & IGNORED \\
2 & TIMEOUT & DISCARDED \\
1 & TRACK & SENT \\
42 & UNINTERESTING & IGNORED \\

postfix\_action IS USED IN A FEW ACTIONS, BUT IT COULD PROBABLY BE EASILY
REMOVED, OR MAYBE I SHOULD KEEP TRACK OF actions INSTEAD\@?

XXX IF I KEEP THESE I NEED TO EXPLAIN WHY I HAVE THEM\@.

    \item [EXPIRED] The mail has been in the queue for five days without
        successful delivery.  A bounce mail will be generated and sent to
        the sender address.

@@

\Figureref{flow chart image} shows the most common paths the data
representing a mail can take through the parser; the complications
described in \sectionref{complications} are excluded for the sake of
clarity.

@@

@@

Everything starts off with a mail entering the system, whether by local
submission via \daemon{postdrop}, by \gls{SMTP}, by re-injection due to
forwarding, or internally generated by Postfix.  Local submission is the
simplest case: a queueid is assigned immediately and the sender address is
logged (action: PICKUP\@; flowchart:~2).  Re-injection due to forwarding
sadly lacks explicit log lines of its own; it is explained in
\sectionref{Re-injected mails}.

\gls{SMTP} is more complicated than local submission:

@@

@@

@@

@@

The crux of this complication is that re-injected mails appear in the log
files without explicit logging indicating their source.  There are two
implicit indications:

@@

    \item The indicator which more commonly introduces re-injection is when
        \daemon{qmgr} selects a mail with a previously unseen queueid for
        delivery, in which case a new data structure will be created for
        that mail.  The mail will be flagged as having unknown origins;
        this flag will subsequently be cleared once the origin has been
        established.  This may also be an indicator that the mail is a
        bounce notification, see
        \sectionref{identifying-bounce-notifications} for details.

    \item Local delivery re-injects the mail and logs a relayed delivery
        rather than delivering directly to a mailbox or program as it
        usually would.\footnote{Relayed delivery is performed by the
        \gls{SMTP} client; local delivery means local to the server, i.e.\
        an address the server is final destination for.}  In this case the
        mail may already have been created (described above) and the
        unknown origin flag will be cleared; if not a new data structure
        will be created.  In both cases the re-injected mail is marked as a
        child of the original mail.  The log line in question is:

        This always occurs for re-injected mail but typically occurs after
        the first indicator.  This log line is required to tie the parent
        and child mails together and so is central to the process of
        tracking re-injected mails.

@@

    \item If the mail is of unknown origin it is assumed to be a child mail
        whose parent has not yet been identified.  Mark the mail as ready
        for entry in the database and wait for the parent to deal with it.
        The mail should not have any subsequent log lines; only its parent
        should refer to it.

    \item If the mail is a child mail then it has already been tracked: as
        with all other mail, the data is cleaned up, the child is entered
        in the database, and then deleted from the state tables.  The child
        mail will be removed from the parent mail's list of children; if
        this is the last child and the parent has already been entered in
        the database, the parent will also be deleted from the state
        tables.

    \item The last alternative is that the mail is a parent mail.
        Regardless of the state of its children its data is cleaned up and
        entered in the database.  The parent may have children that are
        waiting to be entered in the database; the data for each of those
        children is cleaned up and entered in the database, then deleted
        from the state tables.  The parent may also have outstanding
        children which are not yet delivered, in which case the parent must
        be retained in the state tables until those children are finished
        with.  As soon as the last child is deleted from the state tables
        the parent will also be deleted from the state tables.
